Mechanising Hylomorphisms for Extraction of Executable Code

### 2025-03-24 paper/paper.tex
edits for sections 1 and 2

Hylomorphisms are very powerful non-structural recursive algorithms that can be
used to implement structural recursion, as well as many other known recursion
schemes such as mutual recursion, accumulators, primitive recursion and
course-of-values iteration (dynamic programming)~\cite{HinzeWG15}. This power
comes at a price. Since hylomorphisms are not necessarily structural, extra-care
is needed when encoding them in a type theory. In particular, the divide part of
the algorithm must be \emph{recursive} that is the input needs to be split into
smaller parts (not necessarily following the structure of the input data).
On the other hand, a key benefit is the reusability of termination proofs: once
the correctness of the divide phase is established, the same proof can be reused
for any other conquer phase we choose. Moreover, in the case of structural
recursion, the divide phase naturally follows the structure of the data, making
it correct by definition and eliminating the need for additional proofs.

Implementations of recursion schemes currently focus on non-terminating
languages (e.g. Haskell) and they would benefit from a type-theoretical
formalisation which leverages the underlying prover's logic to ensure
correctness of definitions.  While existing mechanisations do exist they either
do not cover a large subset of the recursion schemes or are not suitable for
program calculation or code extraction (see Section~\ref{sec:related-work},
Related Work).

In this work we mechanise hylomorphisms, thus reaping the benefits of their
generality and their algebraic properties. We do this in a way which allows the
user to write specifications of their code, reason about program optimisations,
and eventually extract humanly readable functional code which can be run. Our
extracted code does not contain unsafe casts. We focus strongly on avoiding
unsafe casts because, even if the extracted code has been verified, simple
interoperations with them can lead to incorrect behaviour or even segmentation
faults~\cite{forster:hal-04329663} and, moreover, it invalidates the
fast-and-loose principle~\cite{DanielssonHJG06}.

@@

Furthermore folds and unfolds can be shown to capture a wide
range of recursion schemes such as primitive recursion, mutual recursion,
dynamic programming algorithms, polymorphic recursion, recursion with
accumulators and so on. However, for divide-and-conquer algorithms folds need to
be generalised to hylomorphisms which provide the ultimate basic building block
for any other recursion scheme. This is done by looking at recursion schemes
from the point of view of category theory.

>>>>

Hylomorphisms are highly versatile, non-structural recursive algorithms that are
capable of implementing structural recursion as well as other known recursion
schemes like mutual recursion, accumulators, primitive recursios, and
course-of-values iteration (dynamic programming)~\cite{HinzeWG15}. However, this
versatility comes with a trade-off: because hylomorphisms are not inherently
structurally recursive, special attention is required when encoding them in a
type theory. Specifically, the ``divide'' phase of the algorithm must be
recursive, meaning the input needs to be broken down into smaller parts, which
do not necessarily align with the structure of the input data. On the other
hand, a key benefit is the reusability of termination proofs: once the
correctness of the divide phase is established, the same proof can be reused for
any other conquer phase we choose. Moreover, in the case of structural
recursion, the divide phase naturally follows the structure of the data, making
it correct by definition and eliminating the need for additional proofs.

Implementations of recursion schemes generally focus on non-terminating
languages (e.g. Haskell) and they would benefit from a type-theoretical
formalisation which leverages the underlying prover's logic to ensure
correctness of definitions.  While existing mechanisations do exist they either
do not cover a large subset of the recursion schemes or are not suitable for
program calculation or code extraction (see Section~\ref{sec:related-work},
Related Work).

In this work, we mechanise hylomorphisms, reaping the benefits of their
generality and algebraic properties. Our approach enables users to write
high-level specifications for their programs, reason about program
optimisations, and ultimately extract human-readable, executable functional
code. The extracted code is free from unsafe casts, a priority in our design. We
focus strongly on avoiding unsafe casts because, even if the extracted code has
been verified, simple interoperations with them can lead to incorrect behaviour
or even segmentation faults~\cite{forster:hal-04329663} and, moreover, it
invalidates the fast-and-loose principle~\cite{DanielssonHJG06}.

@@

Furthermore, folds and unfolds can be shown to capture a wide range of recursion
schemes such as primitive recursion, mutual recursion, dynamic programming
algorithms, polymorphic recursion, and recursion with accumulators. However, for
divide-and-conquer algorithms, folds need to be generalised to hylomorphisms,
which provide the ultimate basic building block for any other recursion scheme.
To do this, we look at recursion schemes from the point of view of category
theory.
